BHPWAVE: An adiabatic gravitational waveform model for compact objects undergoing quasi-circular inspirals into rotating massive black holes

### 2023-10-30 ms.tex

Extreme-mass-ratio inspirals (EMRIs) are binaries composed of a compact object with mass $\mu \sim 10 M_\odot$ inspiraling into a massive black hole with mass $M \sim 10^6 M_\odot$. They emit gravitational waves in the millihertz (mHz) band for months to years, making them promising sources for future space-based detectors, such as the Laser Interferometer Space Antenna (LISA) \cite{NASA11, ESA12}. The prolonged evolution and rich harmonic structure of an EMRI waveform communicates a wealth of information about the binary. Thus, we expect LISA to measure the masses, spins, and orbital characteristics of observed EMRIs with unprecedented accuracy \cite{BabaETC17}. These measurements will provide novel observations of massive black holes and their surrounding environments, while also facilitating high-precision tests of general relativity \cite{BerrETC19}. Extracting this information from an observed signal, however, will likely require subradian-accurate models of EMRI gravitational wave emission.

@@

Nonetheless purely adiabatic models---which only include $\Phi_\mathrm{0PA}$ in the gravitational wave phasing---may be suitable for detecting (and possibly characterizing) EMRI signals and are invaluable tools for developing data analysis pipelines for EMRI search and parametrization. They capture almost all of the phasing information and relativistic behavior of EMRI signals, making them powerful probes of astrophysical parameter space. The theoretical foundation and numerical calculation of adiabatic waveforms is also well understood, and has been performed for eccentric, precessing EMRIs \cite{HughETC21}.

@@

Additionally, \texttt{bhpwave} does not provide the first adiabatic calculation of quasi-circular inspirals. These systems have long been studied in the literature \cite{Detw78, KennOri96, FinnThor00, Hugh00b, GralHughWarb16}, and other works have even modeled eccentric, equatorial and eccentric, inclined (precessing) EMRIs for a restricted number massive black hole spins \cite{FujiShib20, SkouLuke22, HughETC21}. Recent works have also produced waveforms that include the effects due to the small body's spin \cite{SkouLuke22, DrumETC23}. However, as aforementioned, it is challenging to implement an adiabatic model that is efficient, accurate, and accessible across a majority of the parameter space, especially for arbitrary values of the massive black hole spin. Even when neglecting eccentricity, the inclusion of massive black hole spin presents a number of challenges that are less pronounced for binaries with non-spinning bodies, such as the larger range of radial separations accessible to spinning binaries, the faster frequency evolution for more deeply bound orbits, and the higher harmonic content needed for modeling rapidly-rotating systems. By building upon the theoretical foundations described in the literature, our aim for \texttt{bhpwave} is to provide an efficient and accessible model that fills in a new area of parameter space and provides an important stepping stone for developing more advanced waveform codes that incorporate all possible orbital effects.

@@

The amplitudes display a number of interesting properties, particularly as we move into the near-horizon extremal Kerr (NHEK) regime, i.e., $1-\hat{a}^2 \ll 1$ and $(r-r_+)/r_+ \ll 1$. For $\hat{a} \lesssim 0.995$ the amplitudes peak at the ISCO and decay with the orbital frequency, but for $\hat{a} \gtrsim 0.995$ the amplitudes can peak well before reaching the ISCO. This behavior has been reported in previous investigations of extremal Kerr black holes \cite{GralPorfWarb15, GralHughWarb16}. For the dominant $\ell = m = 2$ mode, the waveform amplitudes reach their maximum magnitudes near the stationary surface of the outer ergosphere $r_E^+ = 2M$, as demonstrated in the top panel of Figure \ref{fig:amplitudeVariation}. The modes do not have the same turnover behavior for $\ell \neq m$, but instead plateau around $r_E^+$ before rising again as they approach the ISCO. This is due to our reprojection of the $(j, m)$-modes of a spin-weighted spheroidal harmonic basis to the $(\ell, m)$-modes of a spherical basis, given in \eqref{eqn:XfromR}. As demonstrated in \cite{GralPorfWarb15}, larger spheroidal $j$ modes (for a fixed $m$) will decay more rapidly as one approaches the ISCO in the NHEK regime. Therefore, in the spherical $(\ell, m)$-basis, the near-ISCO behavior is dominated by the contribution from the modes around $j=m$, even when $\ell \gg m$. Furthermore, for $\ell > m$ the spherical-spheroidal mixing with many $j\sim m$ modes prevents the turnover and decay of $A_{\ell m}$ as $\Omega \rightarrow \Omega_\mathrm{ISCO}$.

Finally, we consider the relative power between the $\ell = m = 2$ and $\ell = m = 20$ modes for a given orbital frequency, which we approximate with the ratio $|A_{20,20}/A_{2,2}|^2$. In the bottom panel of Figure \ref{fig:amplitudeVariation}, we plot this ratio as a function of $r_0$ for different values of $\hat{a}$. We see that  $|A_{20,20}/A_{2,2}|^2 < 10^{-4}$ for all orbital frequencies and spins, placing a conservative upper bound for the relative power contributed by the $\ell = m = 20$ mode at a fixed frequency. For realistic EMRIs, the relative \emph{total} power (i.e., the power integrated across the entire frequency evolution of an inspiral) contributed by the $\ell = m = 20$ mode is $\lesssim 10^{-7}$, as we demonstrate in the following section. Therefore, we find it unnecessary to include higher harmonic contributions beyond $\ell = m = 20$, even for rapidly-spinning black holes.

@@

\begin{figure*}[!htp]
    \centering
    \includegraphics[width=0.98\linewidth]{figures/few_bhpwave_overlay.pdf}
    \includegraphics[width=0.98\linewidth]{figures/few_phase_comparison.pdf}
    \caption{Comparisons between waveforms produced by the FEW and \texttt{bhpwave} models for an EMRI with intrinsic parameters $(M, \mu, a, p_0, e_0, x) = (10^6 M_\odot, 10 M_\odot, 0, 12.05, 0, 1)$. The top panels show the evolution of the strain $h_+$ with time $t$. The solid (blue) line corresponds to the waveform computed by FEW, while the dashed (orange) line is produced by \texttt{bhpwave}. The three different top panels focus on three small time windows within the full four year signal. The bottom panel plots the difference in the gravitational wave phases between the waveform models. The solid (cyan) line shows the phase difference when the waveforms are evaluated at the same set of parameters. The dashed (black) line is the phase difference after rescaling the FEW parameters to account for different definitions of $GM_\odot$.}
    \label{fig:compareFEW}
\end{figure*}

In the final column of Table \ref{tab:modeCount} we also report the relative power of the $\ell = m = 20$ mode, $P_{20,20}/P_\mathrm{tot}$, where $P_\mathrm{tot}$ is the summed power from all selected $(\ell, m)$-modes. Consequently, this mode is only included if $\epsilon_\mathrm{mode} \leq P_{20,20}/P_\mathrm{tot}$. Previous investigations of adiabatic model suggest that $ 10^{-2} \lesssim \epsilon_\mathrm{mode} \lesssim 10^{-5}$ is a sufficient threshold range to prevent systematic biases in EMRI data analysis for LISA \cite{KatzETC20}. Thus, we see that there is no need to go beyond the $\ell = 20$ modes for the range of spin values considered by our model.

\renewcommand{\arraystretch}{1.5}
\begin{table}
    \caption{The number of selected $(\ell, m)$-modes $N_\mathrm{mode}$ for a given inspiral $(\hat{a}, M, \mu, T)$ and mode selection threshold $\epsilon_\mathrm{mode}$. The initial conditions are chosen so that each system reaches the ISCO after $T$ years. In the final column we also include the relative power in the $\ell = m = 20$ mode, $P_{20,20}/P_\mathrm{tot}$, where $P_\mathrm{tot}$ is the summed power from all selected $(\ell, m)$-modes.}
    \label{tab:modeCount}
    \centering
    \begin{tabular*}{\linewidth}{c @{\extracolsep{\fill}} cccccc}
        \hline
        \hline
         $\hat{a}$ & $M \,(M_\odot)$ & $\mu \,(M_\odot)$ & $T$ (yrs) & $\epsilon_\mathrm{mode}$ & $N_\mathrm{mode}$ & $\frac{P_{20,20}}{P_\mathrm{tot}}$ 
         \\
         \hline
         $0.5000$ & $10^6$ & $10$ & $0.1$ & $10^{-5}$ & 14 & $5\times10^{-12}$
         \\
         $0.9000$ & $10^6$ & $10$ & $0.1$ & $10^{-5}$ & 18 & $1\times10^{-9\phantom{0}}$
         \\
         $0.9950$ & $10^6$ & $10$ & $0.1$ & $10^{-5}$ & 24 & $1\times10^{-7\phantom{0}}$
         \\
         \hline
         $0.9999$ & $10^5$ & $10$ & $0.1$ & $10^{-5}$ & 17 & $7\times10^{-8\phantom{0}}$
         \\
         $0.9999$ & $10^6$ & $10$ & $0.1$ & $10^{-5}$ & 28 & $1\times10^{-6\phantom{0}}$
         \\
         $0.9999$ & $10^7$ & $10$ & $0.1$ & $10^{-5}$ & 39 & $2\times10^{-5\phantom{0}}$
         \\
         \hline
         $0.9999$ & $10^6$ & $10$ & $0.5$ & $10^{-5}$ & 23 & $4\times10^{-7\phantom{0}}$
         \\
         $0.9999$ & $10^6$ & $10$ & $2.0$ & $10^{-5}$ & 20 & $2\times10^{-7\phantom{0}}$
         \\
         $0.9999$ & $10^6$ & $10$ & $4.0$ & $10^{-5}$ & 18 & $1\times10^{-7\phantom{0}}$
         \\
         \hline
         $0.9999$ & $10^6$ & $10$ & $0.1$ & $10^{-3}$ & 10 & $1\times10^{-6\phantom{0}}$
         \\
         $0.9999$ & $10^6$ & $10$ & $0.1$ & $10^{-4}$ & 16 & $1\times10^{-6\phantom{0}}$
         \\
         $0.9999$ & $10^6$ & $10$ & $0.1$ & $10^{-6}$ & 40 & $1\times10^{-6\phantom{0}}$
         \\
         \hline
         \hline
    \end{tabular*}
\end{table}

@@

\begin{figure*}[bhtp]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/flux_comparison.pdf}
    \caption{Fractional error between our interpolated flux data $\mathcal{F}_E^I$ and the flux data reported within the Circular Orbit Self-force Data repository in the BHPToolkit. The left panel displays errors across the entire domain covered by our spline. The middle plot focuses on a region where we find strong disagreement between our model and the repository. The right plot compares our data to flux calculations from a high-precision (>100 digits) Mathematica code in this problem region, demonstrating that our results are consistent with high-precision calculations.}
    \label{fig:fluxComparison}
\end{figure*}

>>>>

Extreme-mass-ratio inspirals (EMRIs) are binaries composed of a compact object with mass $\mu \sim 10 M_\odot$ inspiraling into a massive black hole with mass $M \sim 10^6 M_\odot$. They emit gravitational waves in the milliHertz (mHz) band for months to years, making them promising sources for future space-based detectors, such as the Laser Interferometer Space Antenna (LISA) \cite{NASA11, ESA12}. The prolonged evolution and rich harmonic structure of an EMRI waveform communicates a wealth of information about the binary. Thus, we expect LISA to measure the masses, spins, and orbital characteristics of observed EMRIs with unprecedented accuracy \cite{BabaETC17}. These measurements will provide novel observations of massive black holes and their surrounding environments, while also facilitating high-precision tests of general relativity \cite{BerrETC19}. Extracting this information from an observed signal, however, will likely require subradian-accurate models of EMRI gravitational wave emission.

@@

Nonetheless purely adiabatic models---which only include $\Phi_\mathrm{0PA}$ in the gravitational wave phasing---may be suitable for detecting (and possibly characterizing) EMRI signals \cite{GairJone06} and are invaluable tools for developing data analysis pipelines for EMRI search and parametrization. They capture almost all of the phasing information and relativistic behavior of EMRI signals, making them powerful probes of astrophysical parameter space. The theoretical foundation and numerical calculation of adiabatic waveforms is also well understood, and has been performed for eccentric, precessing EMRIs \cite{HughETC21}.

@@

It is important to note that \texttt{bhpwave} does not provide the first adiabatic calculation of quasi-circular inspirals. These systems have long been studied in the literature \cite{Detw78, KennOri96, FinnThor00, Hugh00b, GralHughWarb16}, and other works have even modeled eccentric, equatorial and eccentric, inclined (precessing) EMRIs for a variety of massive black hole spins \cite{FujiShib20, SkouLuke22, HughETC21}. Recent works have also produced waveforms that include the effects due to the small body's spin \cite{SkouLuke22, DrumETC23}. However, as aforementioned, it is challenging to implement an adiabatic model that is efficient, accurate, and accessible across a majority of the parameter space, especially for arbitrary values of the massive black hole spin. Even when neglecting eccentricity, the inclusion of spin presents a number of challenges that are less pronounced for binaries with non-spinning bodies, such as the larger range of radial separations accessible to spinning binaries, the faster frequency evolution for more deeply bound orbits, and the higher harmonic content needed for modeling rapidly-rotating systems. By building upon the theoretical foundations described in the literature, our aim for \texttt{bhpwave} is to provide an efficient and accessible model that fills in a new area of parameter space and provides an important stepping stone for developing more advanced waveform codes that incorporate all possible orbital effects.

@@

The amplitudes display a number of interesting properties, particularly as we move into the near-horizon extremal Kerr (NHEK) regime, i.e., $1-\hat{a}^2 \ll 1$ and $(r-r_+)/r_+ \ll 1$. For $\hat{a} \lesssim 0.995$ the amplitudes peak at the ISCO and decay with the orbital frequency, but for $\hat{a} \gtrsim 0.995$ the amplitudes can peak well before reaching the ISCO. This behavior has been reported in previous investigations of extremal Kerr black holes \cite{GralPorfWarb15, GralHughWarb16}. For the dominant $\ell = m = 2$ mode, the waveform amplitudes reach their maximum magnitudes near the stationary surface of the outer ergosphere $r_E^+ = 2M$, as demonstrated in the top panel of Figure \ref{fig:amplitudeVariation}. The modes do not have the same turnover behavior for $\ell \neq m$, but instead plateau around $r_E^+$ before rising again as they approach the ISCO, which is highlighted in the middle panel of \ref{fig:amplitudeVariation} for all $m =2$ modes. This behavior is due to our reprojection of the $(j, m)$-modes of a spin-weighted spheroidal harmonic basis to the $(\ell, m)$-modes of a spherical basis, given in \eqref{eqn:XfromR}. As demonstrated in \cite{GralPorfWarb15}, larger spheroidal $j$ modes (for a fixed $m$) will decay more rapidly as one approaches the ISCO in the NHEK regime. Therefore, in the spherical $(\ell, m)$-basis, the near-ISCO behavior is dominated by the contribution from the modes around $j=m$, even when $\ell \gg m$. Furthermore, for $\ell > m$ the spherical-spheroidal mixing with many $j\sim m$ modes prevents the turnover and decay of $A_{\ell m}$ as $\Omega \rightarrow \Omega_\mathrm{ISCO}$.

Finally, we consider the relative power between the $\ell = m = 2$ and $\ell = m = 15$ modes for a given orbital frequency. In the bottom panel of Figure \ref{fig:amplitudeVariation}, we plot the $A_{\ell, m = \ell}$ amplitudes as functions of $r_0$ for different values of $\hat{a}$. We see that $|A_{15,15}/A_{2,2}|^2 \lesssim 5\times 10^{-4}$ for all orbital frequencies and spins, placing a conservative upper bound for the relative power contributed by the $\ell = m = 15$ mode at a fixed frequency. For most year-long EMRI signals, the relative \emph{total} power (i.e., the power integrated across the entire frequency evolution of an inspiral) contributed by the $\ell = m = 15$ mode is $\lesssim 10^{-6}$, as we demonstrate in the following section. Therefore, we find it unnecessary to include higher harmonic contributions beyond $\ell = m = 20$, even for rapidly-spinning black holes.

@@

\begin{figure*}[!htp]
    \centering
    \includegraphics[width=0.98\linewidth]{figures/few_bhpwave_overlay.pdf}
    \includegraphics[width=0.98\linewidth]{figures/few_phase_comparison.pdf}
    \caption{Comparisons between waveforms produced by the FEW and \texttt{bhpwave} models for an EMRI with intrinsic parameters $(M, \mu, a, p_0, e_0, x) = (10^6 M_\odot, 10 M_\odot, 0, 12.05, 0, 1)$. The top panels show the evolution of the strain $h_+$ with time $t$. In the top plots, the solid (blue) lines correspond to the waveform computed by FEW, while the dashed (orange) lines are produced by \texttt{bhpwave}. The three different panels focus on three small time windows within the full four year signal. The bottom panel plots the difference in the gravitational wave phases between the waveform models. The solid (cyan) line shows the phase difference when the waveforms are evaluated at the same set of parameters. The dashed (black) line is the phase difference after rescaling the FEW parameters to account for different definitions of $GM_\odot$.}
    \label{fig:compareFEW}
\end{figure*}

In the final column of Table \ref{tab:modeCount} we also report the relative power of the $\ell = m = 15$ mode, $P_{15,15}/P_\mathrm{tot}$, where $P_\mathrm{tot}$ is the summed power from all selected $(\ell, m)$-modes. Consequently, this mode is only included if $\epsilon_\mathrm{mode} \leq P_{15,15}/P_\mathrm{tot}$. Previous investigations of adiabatic model suggest that $ 10^{-2} \lesssim \epsilon_\mathrm{mode} \lesssim 10^{-5}$ is a sufficient threshold range to prevent systematic biases in EMRI data analysis for LISA \cite{KatzETC20}. Thus, we see that there is no need to go beyond the $\ell = 15$ modes for the range of spin values considered by our model.

\renewcommand{\arraystretch}{1.5}
\begin{table}
    \caption{The number of selected $(\ell, m)$-modes $N_\mathrm{mode}$ for a given inspiral $(\hat{a}, M, \mu, T)$ and mode selection threshold $\epsilon_\mathrm{mode}$. The initial conditions are chosen so that each system reaches the ISCO after $T$ years. In the final column we also include the relative power in the $\ell = m = 15$ mode, $P_{15,15}/P_\mathrm{tot}$, where $P_\mathrm{tot}$ is the summed power from all selected $(\ell, m)$-modes.}
    \label{tab:modeCount}
    \centering
    \begin{tabular*}{\linewidth}{c @{\extracolsep{\fill}} cccccc}
        \hline
        \hline
         $\hat{a}$ & $M \,(M_\odot)$ & $\mu \,(M_\odot)$ & $T$ (yrs) & $\epsilon_\mathrm{mode}$ & $N_\mathrm{mode}$ & $\frac{P_{15,15}}{P_\mathrm{tot}}$ 
         \\
         \hline
         $0.5000$ & $10^6$ & $10$ & $0.1$ & $10^{-5}$ & 14 & $2\times10^{-9}$
         \\
         $0.9000$ & $10^6$ & $10$ & $0.1$ & $10^{-5}$ & 18 & $1\times10^{-7}$
         \\
         $0.9950$ & $10^6$ & $10$ & $0.1$ & $10^{-5}$ & 24 & $3\times10^{-6}$
         \\
         \hline
         $0.9999$ & $10^5$ & $10$ & $0.1$ & $10^{-5}$ & 17 & $8\times10^{-7}$
         \\
         $0.9999$ & $10^6$ & $10$ & $0.1$ & $10^{-5}$ & 28 & $1\times10^{-5}$
         \\
         $0.9999$ & $10^7$ & $10$ & $0.1$ & $10^{-5}$ & 39 & $2\times10^{-4}$
         \\
         \hline
         $0.9500$ & $10^5$ & $10$ & $1.0$ & $10^{-5}$ & 11 & $5\times10^{-9}$
         \\
         $0.9500$ & $10^6$ & $10$ & $1.0$ & $10^{-5}$ & 15 & $6\times10^{-8}$
         \\
         $0.9500$ & $10^7$ & $10$ & $1.0$ & $10^{-5}$ & 23 & $1\times10^{-6}$
         \\
         \hline
         $0.9999$ & $10^6$ & $10$ & $0.5$ & $10^{-5}$ & 23 & $4\times10^{-6}$
         \\
         $0.9999$ & $10^6$ & $10$ & $2.0$ & $10^{-5}$ & 20 & $2\times10^{-6}$
         \\
         $0.9999$ & $10^6$ & $10$ & $4.0$ & $10^{-5}$ & 18 & $1\times10^{-6}$
         \\
         \hline
         $0.9999$ & $10^6$ & $10$ & $0.1$ & $10^{-3}$ & 10 & $1\times10^{-5}$
         \\
         $0.9999$ & $10^6$ & $10$ & $0.1$ & $10^{-4}$ & 16 & $1\times10^{-5}$
         \\
         $0.9999$ & $10^6$ & $10$ & $0.1$ & $10^{-6}$ & 40 & $1\times10^{-5}$
         \\
         \hline
         \hline
    \end{tabular*}
\end{table}

@@

\begin{figure*}[bhtp]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/flux_comparison.pdf}
    \caption{Fractional error between our interpolated flux data $\mathcal{F}_E^I$ and the flux data reported within the Circular Orbit Self-force Data repository in the BHPToolkit. The left panel displays errors across the entire domain covered by our spline. The middle plot focuses on a region where we find strong disagreement between our model and the repository. The right plot compares our data to flux calculations from a high-precision ($>100$ digits) Mathematica code in this problem region, demonstrating that our results are consistent with high-precision calculations.}
    \label{fig:fluxComparison}
\end{figure*}
